---
title: "Origin of coexisting species"
author: "William Ou"
date: "2024-09-23"
output: pdf_document
---

# Model description

## Fitness function:

$$
\tag{1}
W_i = \frac{\lambda \psi(z_i, z_j)}{1 + \frac{C(z_i, z_j)}{K(z_i)}} (1-I_i)
$$

-   Fitness of individual $i$ is the product of its fertility $\lambda$, reduction in fitness due to incompatability of its parents $I_i$, and its ability to find a mate $\psi (z_i, z_j)$ divided by the amount of competition the individual experiences $1+\frac{C(z_i, z_j)}{K(z_i)}$

-   For simplicity, we assume $\lambda$ is equal for all individuals hence the removal of the subscript

## Incompatability:

$$
\tag2{}
I_i = 1 - (1 - \phi)^{n_i}
$$

-   The fitness effect due to the incomptability of individual $i$'s parents ($I_i$) increases geometrically with the number of incompatible BDMI loci $n_i$ between the parents by the factor $1 -\phi$

    ```{r, echo = FALSE, message = FALSE, warning = FALSE, out.width="50%", fig.align = 'center'}
    library(ggplot2)
    library(dplyr)
    expand.grid(n = 0:5, phi = seq(from = 0.1, to = 0.3, by = 0.1))%>%
      mutate(I = 1 - (1 - phi)^n)%>%
      mutate(phi = as.factor(phi))%>%
      ggplot(aes(x = n, y = I, group = phi, col = phi))+ 
      geom_line(size = 1.5)+
      scale_color_manual(values = c('#d0d1e6', '#67a9cf', '#016450'),
                         name = expression(phi))+
      labs(x = expression(n[i]), y = expression(I[i]))+
      theme_classic()+
      theme(axis.title = element_text(size = 18),
            legend.text = element_text(size=15),
            legend.title = element_text(size = 17),
            axis.text = element_text(size = 15))
    ```

-   Individuals are haploids with $B = 5$ unlinked BDMI loci, where each locus has 2 alleles (1 or 0). Number of incompatible BDMI loci $n_i$ is computed as the number of BDMIs of mated pairs that are not the same.

    ```{r, echo = FALSE, message = FALSE, warning = FALSE, out.width="60%", fig.align = 'center'}
    data.frame(male = c(1,1,1,0,1),
               female = c(1,1, 0, 0, 0),
               offspring = c(1,1,1,0,0))%>%
      tibble::rowid_to_column(var = 'locus')%>%
      tidyr::pivot_longer(cols = 2:4, names_to = 'group', values_to = 'state')%>%
      mutate(ypos = plyr::mapvalues(group, from = c('male','female','offspring'), to = c(1.5, 1, 0.3)))%>%
      mutate(ypos = as.numeric(ypos))%>%
      mutate(state = as.factor(state))%>%
      ggplot(aes(x = locus, y = ypos))+
      geom_point(size = 10, aes(fill = state), shape = 21, stroke = 2)+
      scale_fill_manual(values = c('black', 'white'), name = 'Allele')+
      lims(y = c(0, 2), x = c(-0.5,6))+
      annotate('text', x = 1:5, y = 1.8, label = paste0('BDMI', 1:5), size = 4)+
      annotate('text', x = 0, y = c(1, 1.5), label = c('Parent i', 'Parent j'), size = 5)+
      annotate('segment', x = 1:5, xend = 1:5, y = 0.75, yend = 0.5, arrow = arrow(type = "closed", length = unit(0.3, 'cm')))+
      annotate('text', x = 0, y = 0.3, label = 'Offspring', size = 5)+
      theme_void()
    ```

## Mating probability

-   The probability of female $i$ mating with male $j$ (e.g., $\psi (z_i, z_j)$) is determined by the distance between their ecological trait $z$ and the strength of assortative mating ($\sigma_m$) normalized over potential male mates of female $i$ (i.e., all females mate with probability = 1 given that males are present) according to the equation

$$
\tag{3}
\psi (z_i, z_j)  = \frac{e^{\frac{(z_i - z_j)^2}{2\sigma_m^2}}} {\sum_j{e^{\frac{(z_i - z_j)^2}{2\sigma_m^2}}}}
$$

-   This model thus incorporates mate competition such that at each time step, some males may not mate, while some may mate with multiple females

-   **NOT YET INCORPORATED:** $\sigma_m$ currently fixed across individuals

## Resource competition

-   Individuals in a population compete for resources according to their differences in ecological trait $z_i$ (e.g., $C_i(z_i, z_j)$) scaled by resources available to the individual $K(z_i)$ (i.e., $C(z_i, z_j) / K(z_i)$)

-   Specifically, $C(z_i, z_j)$ is the competition experienced by individual $i$ in the presence of individual $j$s defined as,

$$
\tag{4}
C_{ij} = exp\frac{-\sum_{i \neq j}(z_i - z_j)^2}{2\sigma_z^2}$$

where $\sigma_z^2$ is the competition kernel (or niche width) which we assume is fixed and equal for all individuals

-   In the absence of competitors, an individual with phenotype $z_i$ has access to $K(z_i)$ amount of resources which is defined as,

$$
\tag{5}
K(z_i) =   K_oexp(\frac{-(z_i - \theta_p)^2}{2\sigma_k^2})
$$

where $K_o$ is the maximum amount of resources, or the resources available to an individual with an optimum phenotype for that patch, $\theta_p$, and $\sigma_k^2$ determines the rate in which resource availability decays with the distances between $z_i$ and the optimum (i.e., strength of stabilizing selection).

## Inheritance and mutations

-   **BDMIs**: Offsprings inherit each BDMI locus independently via random sampling of parent alleles. Each inherited allele has a probability of flipping $\mu_i$

-   **Assortative mating** $\sigma_m$: *currently not incorporated yet*

-   **Ecological trait** $z_i$: offspring $z$ is drawn from a Gaussian distribution with mean equal the average of parent $z_i$ and variance, $\mu_z$ (i.e., $\mathcal{N}(z_{parents}, \mu_z)$)

-   **Location** $p_i$: offspring are born into the patch of its parents with probability $1 - m$

```{r, echo = FALSE,message=FALSE, warning=FALSE}
# Load required packages
library(knitr)
library(kableExtra)

# Create the data frame for the table
table_data <- data.frame(
`Model parameter` = c("$\\lambda$", "$\\phi$", "$\\sigma_m$", "$\\sigma_z$", "$K_o$", "$\\theta_p$"),
Description = c("Fertility", "Incompatability", "Assortative variance", "Niche width", "Maximum resource availability", "Optimum of patch p"),
Value = c(5, 0.1, 0.05, 0.05, 30, 0))

# Bold the column names using LaTeX formatting
colnames(table_data) <- c("\\textbf{Model parameter}", "\\textbf{Description}", "\\textbf{Value [interval]}")

# Create the table using kable and kableExtra
table_data %>%
  kable(caption = "List of model parmaters and their description", align = 'c', format = "latex", escape = FALSE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## R functions

1.  Initialize initial population `initial_condition(N, bdmi_B)`

-   This function initializes a data frame that into a form that the main simulator function interprets

-   `N` is the arugment specifying the number of individuals to start and `bdmi_B` is the number of BDMI loci each individual has

2.  Simulator `ibm(population, num_gens, assort_sigma = 0.05, phi = 0.2, birth = 10, theta = c(0, 0.01), sig2_alpha = 0.05, K0 = 50, sig2_k = 0.05, migration_p = 0.1, bdmi_mu = 10^-3, eco_mu = 10^-2)`

-   This is the main function that implments the individual-based model

-   `population` is a dataframe containing the initial condition of the models (output of `initial_condition()`

-   `num_gens` specifies the number of generations in which to run the model for

-   Other arguments specify the parameter values of the model
